---
title: "Data Cleaning and Initial Plots"
author: "James"
format: html
editor: source
---
Loading packages.
```{r}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(scales)
```
Loading in raw data.
```{r}
raw_data <- read_csv("life-satisfaction-vs-life-expectancy.csv")

```
Fixing issue where region is only reported once per country.
```{r}
region_map <- raw_data %>%
  filter(!is.na(`World regions according to OWID`)) %>%
  distinct(Entity, `World regions according to OWID`) %>%
  deframe() 

filled_region <- raw_data %>%
  mutate(
    `World regions according to OWID` = coalesce(
      `World regions according to OWID`,
      region_map[Entity]
    )
  )
```

Cleaning data to eliminate rows with missing variables for Life Expectancy and Cantril Ladder Score.

Also rename Life Expectancy column.
```{r}
cleaned <- filled_region %>%
  mutate(
    `Life expectancy (years)` = `Life expectancy - Sex: all - Age: 0 - Variant: estimates`
  ) %>%
  select(-`Life expectancy - Sex: all - Age: 0 - Variant: estimates`) %>%
  filter(
    !is.na(`Life expectancy (years)`),
    !is.na(`Cantril ladder score`)
  )
```

Collecting selected countries for initial inspection and plotting
```{r}
selected_countries <- c(
  "Australia", "South Korea", "Sweden", "Germany",
  "United States", "Mexico", "Brazil", "Vietnam", "France"
)

cleaned %>%
  filter(Entity %in% selected_countries) %>%
  write_csv("selected_countries.csv")

initial_countries <- read_csv("selected_countries.csv")

```

Initial plotting
```{r}
ggplot(
  initial_countries %>% filter(Year == 2023),
  aes(
    x     = `Cantril ladder score`,
    y     = `Life expectancy (years)`,
    color = `World regions according to OWID`
  )
) +
  geom_point(size = 4) +
  scale_x_continuous(
    name   = "Life Satisfaction (Cantril ladder score)",
    limits = c(1, 8)
  ) +
  scale_y_continuous(
    name   = "Life Expectancy (years)",
    limits = c(45, 90)
  ) +
  labs(
    title = "Life Satisfaction vs Life Expectancy in 2023",
    color = "Region"
  ) +
  theme_minimal()
```

```{r}
ggplot(
  initial_countries %>%
    filter(Entity == "Australia") %>%
    arrange(Year),
  aes(
    x     = `Cantril ladder score`,
    y     = `Life expectancy (years)`,
    color = Year
  )
) +
  geom_path() +
  geom_point(size = 3) +
  scale_color_viridis_c(
    breaks = pretty_breaks(n = 5)
  ) +
  labs(
    title = "Australia: Life Satisfaction vs Life Expectancy Over Time",
    x     = "Life Satisfaction (Cantril ladder score)",
    y     = "Life Expectancy (years)",
    color = "Year"
  ) +
  theme_minimal()
```

Repeating the above plot in a 2x2 faceted grid.
Note that scales have been allowed to be freely set. Direct comparison of values is not intended in this graph, only a visual inspection of the overall trajectory of each country.

```{r}
initial_countries %>%
  filter(Entity %in% c("Australia", "United States", "Mexico", "Vietnam")) %>%
  arrange(Year) %>%                          
  ggplot(aes(
    x     = `Cantril ladder score`,
    y     = `Life expectancy (years)`,
    color = Year,
    group = Entity
  )) +
  geom_path(size = 0.8) +                   
  geom_point(size = 3) +
  scale_color_viridis_c(
    breaks = pretty_breaks(n = 5)           
  ) +
  facet_wrap(~ Entity, ncol = 2, scales = "free") +
  labs(
    title = "Life Satisfaction vs Life Expectancy Over Time",
    x     = "Life Satisfaction (Cantril ladder score)",
    y     = "Life Expectancy (years)",
    color = "Year"
  ) +
  theme_minimal()
```






