---
title: "ETC5513 Assignment 3"
subtitle: "Life Satisfaction vs. Life Expectancy"
author: "Often imitated, never duplicated  \n TingTing Wu, Yilin Zhang, James Gray"
format: revealjs
theme: serif
output-file: "ETC5513_Assignment3_Slides.html"
---

```{r}
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
library(ggplot2)
library(dplyr)
library(scales)
```

## Table of Contents {.unnumbered}

- Executive Summary  
- Introduction  
- Methodology  
- Global Trends  
- Regional Comparison  
- Australia-Focused Analysis  
- Results  
- Discussion & Conclusion  
- References  

---

## Executive Summary

This report investigates trends in life expectancy and life satisfaction in Europe and Asia from 2011 to 2023. Europe consistently outperformed Asia in the two indicators, and the differences among countries were smaller. However, there have been significant fluctuations and differences in Asian countries. Some countries lag behind in both indicators, while in others, although life expectancy has risen, the happiness index of life lags behind. The research results emphasize the importance of the all-round development that balances longevity and overall health.

---

## Introduction

With advancements in global medicine, public health, and living standards, life expectancy has continued to reach new heights. It is often regarded as a key indicator of people’s well-being. However, the relationship between life expectancy and life satisfaction may vary across countries and regions, influenced by differences in cultural values, economic development, and societal expectations.This report explores the question: Do Europe and Asia experience similar trends in life expectancy and life satisfaction? To address this question, we analyze publicly available data from [Our World in Data](https://ourworldindata.org/grapher/life-satisfaction-vs-life-expectancy), which includes standardized country-level measures of life expectancy and self-reported life satisfaction. Our analysis focuses on Europe and Asia. By cleaning and analyzing data, we discuss the visualization results and key findings.

---

## Methodology

```{r}
#| echo: false
#| message: false
#| warning: false
raw_data <- read_csv("life-satisfaction-vs-life-expectancy.csv")

region_map <- raw_data %>%
  filter(!is.na(`World regions according to OWID`)) %>%
  distinct(Entity, `World regions according to OWID`) %>%
  deframe() 

filled_region <- raw_data %>%
  mutate(`World regions according to OWID` =
           coalesce(`World regions according to OWID`,
                    region_map[Entity]))


cleaned <- filled_region %>%
  mutate(`Life expectancy (years)` = `Life expectancy - Sex: all - Age: 0 - Variant: estimates`) %>%
  select(-`Life expectancy - Sex: all - Age: 0 - Variant: estimates`) %>%
  filter(!is.na(`Life expectancy (years)`), !is.na(`Cantril ladder score`))

cleaned %>%
  filter(`World regions according to OWID` %in% c("Europe", "Asia")) %>%
  write_csv("europe_asia.csv")

europe_asia <- read_csv("europe_asia.csv")
```

```{r}
#| label: fig-table-asia-europe
#| fig-cap: "Europe and Asia filtered dataset"
head(europe_asia)
```

---

```{r}
#| label: fig-dist-asia-europe
#| fig-cap: "Country-Level Distribution Curves of Life Satisfaction and Life Expectancy in Asia and Europe (2023)"
#| fig-pos: "H"
long_data <- europe_asia %>%
  filter(Year == 2023) %>%
  select(`World regions according to OWID`, 
         `Life expectancy (years)`, 
         `Cantril ladder score`) %>%
pivot_longer(
  cols = c(`Life expectancy (years)`, `Cantril ladder score`),
  names_to = "Variable",
  values_to = "Value")
  
ggplot(long_data, aes(x = Value, fill = `World regions according to OWID`)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~ Variable, ncol = 1, scales = "free") +
  labs(title = "Distribution of Life Expectancy and Life Satisfaction (2023)",
    fill = "Region",
    x = NULL,
    y = "Density") +
    theme_minimal()
```

---

```{r}
#| label: fig-dens-asia-europe
#| fig-cap: "Country-Level Density Map of Life Satisfaction and Life Expectancy in Asia and Europe (2023)"
#| fig-pos: "H"
  ggplot(europe_asia %>% filter(Year == 2023), 
         aes(x = `Cantril ladder score`, 
             y = `Life expectancy (years)`)) +
    geom_point(aes(color = `World regions according to OWID`), size = 2) +
    stat_density_2d(aes(fill = `World regions according to OWID`), 
                    geom = "polygon", alpha = 0.2, color = NA, show.legend = FALSE) +
    labs(title = "Density Map for Europe and Asia (2023)",
         x = "Life Satisfaction",
         y = "Life Expectancy") +
    theme_minimal()

```

## Results

```{r}
#| echo: false
# Load filtered dataset that includes only Europe and Asia
europe_asia <- read_csv("europe_asia.csv")

# Compute population-weighted yearly averages for each region
region_avg <- europe_asia %>%
  
  # Remove any missing values in key variables
  filter( 
    !is.na(`Life expectancy (years)`),
    !is.na(`Cantril ladder score`),
    !is.na(`Population (historical)`),
    !is.na(Year)
  ) %>%
  group_by(`World regions according to OWID`, Year) %>%
  summarise( 
    
    # Weighted mean life expectancy using historical population
    `Life expectancy (years)` = weighted.mean(
      `Life expectancy (years)`, `Population (historical)`, na.rm = TRUE),
    
    # Weighted mean life satisfaction using same population weights
    `Cantril ladder score` = weighted.mean(
      `Cantril ladder score`, `Population (historical)`, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r}
#| label: fig-trend-asia-europe
#| fig-cap: "Trajectory of Population-Weighted Life Satisfaction and Life Expectancy in Asia and Europe (2011–2023)"

# Identify the most recent year (2023) for each region to use as label position
label_data <- region_avg %>%
  group_by(`World regions according to OWID`) %>%
  filter(Year == max(Year)) %>%
  ungroup()

# Create line plot showing trend of weighted averages for Asia and Europe
ggplot(region_avg, aes(
  x = `Cantril ladder score`,
  y = `Life expectancy (years)`,
  group = `World regions according to OWID`,
  color = Year
)) +
  geom_path(linewidth = 1.2) +   # Line to show trajectory
  geom_point(size = 2) +         # Dots for each year
  scale_color_viridis_c() +      # Color gradient for year
  geom_text(                     
    data = label_data,           # Add region labels at final year
    aes(label = `World regions according to OWID`),
    color = "black",
    vjust = -1
  ) +
  
  # Add padding on x-axis and y-axis
  scale_x_continuous(expand = expansion(mult = c(0.05, 0.1))) + 
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.1))) + 
  labs(
    title = "Yearly Averages: Life Satisfaction vs Life Expectancy",
    x = "Life Satisfaction (Cantril ladder score)",
    y = "Life Expectancy (years)",
    color = "Year"
  ) +
  scale_color_viridis_c(
  breaks = c(2011, 2014, 2017, 2020, 2023)  # Customize colorbar ticks
) +
  theme_minimal()
```

---

```{r}
#| echo: false
#| label: fig-asia-europe-2011-2023
#| fig-cap: "Life Satisfaction and Life Expectancy in Asia and Europe (2011 vs 2023)"

# Filter dataset to include only the years 2011 and 2023 for comparison
filtered_2years <- europe_asia %>%
  filter(Year %in% c(2011, 2023))

# Create a faceted scatterplot comparing life satisfaction and life expectancy
ggplot(filtered_2years, aes(
  x = `Life expectancy (years)`,
  y = `Cantril ladder score`,
  color = `World regions according to OWID`,
  size = `Population (historical)`
)) +
  geom_point(alpha = 0.6) +          # Draw points with some transparency
  facet_wrap(~Year) +                # Create separate panels for each year
  scale_size(range = c(2, 10)) +     # Control range of point sizes
  scale_color_manual(values = c(     # Assign manual colors to regions
    "Asia" = "#1ABC9C",
    "Europe" = "#9B59B6"
  )) +
  labs(
    title = "Life Satisfaction and Life Expectancy (2011 vs 2023)",
    x = "Life Expectancy (Years)",
    y = "Life Satisfaction",
    color = "Region"
  ) +
  theme_minimal() +
  theme(                                  # Centered bold title
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),  
    plot.margin = margin(10, 10, 10, 10)  # Add margin around plot
  ) +
  guides(size = "none")       # Hide size legend to reduce clutter
```

---

## Discussion & Conclusion

The central question in this report is whether there are similar trends in life expectancy and life satisfaction between Europe and Asia. By analyzing the population-weighted data from 2011 to 2023, we found that the overall quality of life in both regions has improved over time. However, Europe countries have made consistent progress in terms of health and subjective well-being. They are closely clustered together, indicating relatively uniform regional development. In contrast, Asian countries showed greater differences and extreme values observed. Some of these countries performed well in both indicators, while others lagged significantly behind. These phenomena indicate that the differences in medical insurance policies, economic conditions and social support systems behind different regions may affect well-being outcomes. There are mostly high-welfare countries in Europe, which ensures its relatively stable life expectancy and high level of satisfaction. However, in Asia, the levels in various aspects vary greatly and there is certain room for improvement.

---

## Recommendations

Recommendations:

-   Some governments in Asian countries should consider using the people's happiness index as an indicator for policy decisions to ensure that life expectancy and quality of life can be further improved.

-   Future research can expand its scope to include other regions around the world and introduce more differences to explore causal relationships.

-   consider adding other variables such as occupational inequality, income inequality and educational level to help further explain the differences and propose more specific improvement measures.

---

## References

- Our World in Data. (n.d.). *Life satisfaction vs. life expectancy*. https://ourworldindata.org/grapher/life-satisfaction-vs-life-expectancy  

---

## Thank You!

**Questions?**  

Group: *Often imitated, never duplicated*  

- TingTing Wu  
- Yilin Zhang  
- James Gray